---
title: "Bird Study Workshops Report"
author: "Stephen Kaluzny"
date: "`r format(Sys.time(), '%B %d, %Y')`"
format:
  html:
    self-contained: true
    theme: cerulean
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height=7, fig.width=8)
library("flextable")
library("viridis")
library("dplyr")
library("ggplot2")
```

```{r combine_birdsseen, echo=FALSE, include=FALSE}
source(here::here("scripts", "combineBirdsSeen.R"))
```

```{r scouts_by_year, echo=FALSE, include=FALSE}
ScoutsByYear <- read.table(here::here("data", "scoutsByYear.txt"),
  sep=" ", header=TRUE)
ScoutsByYear[["Year"]] <- lubridate::year(as.Date(ScoutsByYear[["ClassDate"]]))
ScoutsByYearAll <- ScoutsByYear
ScoutsByYear <- ScoutsByYear |> filter(Year != 2020)
ScoutsByYear[["fYear"]] <- as.factor(ScoutsByYear[["Year"]])
ScoutsByYear[["fScouts"]] <- as.factor(ScoutsByYear[["Scouts"]])
ScoutsByYear[["oScouts"]] <- ordered(ScoutsByYear[["Scouts"]])
ScoutsByYear <- ScoutsByYear |>
  mutate(Finished = BoysFinished + GirlsFinished) |>
  mutate(PercentCompleted = Finished / Scouts * 100)
```

```{r compute_summary_stats, echo=FALSE}
number_years <- nrow(ScoutsByYearAll)
number_field_years <- nrow(ScoutsByYear)
missing_years <-
  setdiff(min(ScoutsByYearAll[["Year"]]):max(ScoutsByYearAll[["Year"]]),
    ScoutsByYearAll[["Year"]])
total_scouts <- sum(ScoutsByYear$Scouts)
min_scouts <- min(ScoutsByYear$Scouts)
max_scouts <- max(ScoutsByYear$Scouts)
total_completed <- sum(ScoutsByYear$BoysFinished) +
  sum(ScoutsByYear$GirlsFinished)
mean_completed <- mean(ScoutsByYear$PercentCompleted)
first_date <- min(dates_by_year)
second_date <- sort(dates_by_year)[2]
```

## Introduction
Mark Oberle and Stephen Kaluzny have been teaching a workshop for the
Bird Study Merit Badge for Boy Scout Troop 186 in Seattle since 2004.

The workshop is typically held on a Saturday in the Spring.
The morning classroom session is usually held in the
Burke Museum of Natural History and Culture
at the University of Washington.
The afternoon field trip session is held at
the Union Bay Natural Area next to Lake Washington.

After the workshop the scouts need to make and deploy a bird feeder or
water bath to complete the merit badge requirements.
Spreading peanut butter on a conifer cone, rolling it in bird seeds and
hanging the cone in a tree is how the majority of the
scouts fullfill this last requirement.
Despite this easy last requirement,
not all scouts who attend the workshop earn the merit badge.

The workshop has been taught for `r number_years` years
(no workshops were held in `r missing_years`).

The first workshop was on `r format(first_date, format="%B %d, %Y")`.
The next workshop was on `r format(second_date, format="%B %d, %Y")`.
After `r lubridate::year(second_date)`,
the workshops were always held in the Spring.

A virtual workshop was held during the first year of the COVID19 pandemic
(2020).
No field trip session was held that year.
Since this was not a complete workshop, the data from 2020
are left out from the results below.

Over the `r number_years` years of the workshop,
`r total_scouts` scouts have attended,
The maximum attendance for a workshop was `r max_scouts` scouts
and there have been several years with only `r min_scouts` attending.

Of the `r total_scouts` scouts who have attended,
`r total_completed` have completed the merit badge
(a `r round(total_completed / total_scouts * 100, 1)` percent completion rate).

Starting in 2019, girls who had joined Scouts attended the workshop.
So far `r sum(ScoutsByYear$Girls)` girls have attended.

## Scouts

### Scouts attending the workshop.

```{r scouts, echo=FALSE}
ScoutsByYear |> ggplot(aes(x=fYear, y=Scouts)) +
  geom_bar(stat="identity", width=0.3, fill="blue", col="blue") +
  xlab("Year") +
  ggtitle("Number of Scouts by Year") +
  theme_light()
```

### Scouts completing the merit badge.
The mean yearly completion rate is
`r round(mean_completed, 2)` percent.

```{r completing, echo=FALSE, eval=FALSE}
ScoutsByYear |>
  ggplot(aes(x=fYear, y=PercentCompleted, fill=fScouts)) +
    # geom_point() +
    geom_bar(stat="identity", width=0.5) +
#    scale_fill_hue() +
#     scale_fill_viridis(discrete = TRUE) +
    labs(fill="Number of Scouts") +
    xlab("Year") +
    ylab("Percent Completed") +
    ggtitle("Percent Completed the Merit Badge by Year") +
    theme(legend.position = "top")
```

```{r completing2, echo=FALSE, eval=FALSE}
ScoutsByYear |>
  ggplot(aes(x=fYear, y=PercentCompleted, fill=Scouts)) +
    # geom_point() +
    geom_bar(stat="identity", width=0.5) +
#    scale_fill_hue() +
#    scale_fill_viridis(discrete = TRUE) +
    scale_fill_viridis() +
    labs(fill="Number of Scouts") +
    xlab("Year") +
    ylab("Percent Completed") +
    ggtitle("Percent Completed the Merit Badge by Year") +
    theme(legend.position = "top")
```

```{r completing3, echo=FALSE}
ScoutsByYear |>
  # ggplot(aes(x=fYear, y=PercentCompleted, fill=Scouts)) +
  ggplot(aes(x=fYear, y=PercentCompleted)) +
    geom_segment(aes(x=fYear, xend=fYear, y=0, yend=PercentCompleted),
      color="skyblue") +
    geom_point(color="blue", size=4, alpha=0.6) +
    theme_light() +
    theme(
      panel.grid.major.y = element_blank(),
      panel.border = element_blank(),
      axis.ticks.y = element_blank()) +
    # labs(fill="Number of Scouts") +
    xlab("Year") +
    ylab("Percent Completed") +
    ggtitle("Percentage of Scouts Completing the Merit Badge by Year")
```

```{r completing4, echo=FALSE, eval=FALSE}
ScoutsByYear |> mutate(Hundred=100) |>
  ggplot(aes(x=fYear, y=Hundred)) +
    theme_light() +
    geom_bar(stat = "identity") +
    geom_bar(aes(x=fYear, y=PercentCompleted),
      col="skyblue", fill="skyblue", stat = "identity") +
    facet_wrap(~ fYear, scale="free_x")
```

## Birds Seen

One of the merit badge requirements is to create a notebook listing
all the birds species seen on the field trip.
The leaders have recorded all the species seen on all the field trips
over the years.

```{r combine_birds, echo=FALSE}
source(here::here("scripts/combineBirdsSeen.R"))
```

```{r order_species, echo=FALSE}
ordered_species <- scan(here::here("data/ordered_species.txt"), what="", sep="\n")
stopifnot(length(setdiff(birds_by_year[["Species"]], ordered_species)) == 0)
```
### Common Birds
These common birds were seen almost every year on the field trip.
Only birds seen by both scouts and the leaders are considered.

```{r common_birds, echo=FALSE}
n75 <- number_years * 0.75
year_count <- birds_by_year |>
  filter(SeenBy == "Everyone") |> pull(Species) |> table()
common_birds <- sort(year_count[year_count >= n75])
tdf <- data.frame(Species=c(names(common_birds)),
    NumberOfYearsSeen=c(common_birds))
# knitr::kable(tdf,
#   row.names=FALSE, align=c("l", "l"),
#   col.names=c("Species", "Number of Years Seen"))
ft <- flextable(tdf, cwidth=c(3.0, 2.0))
ft <- set_header_labels(ft,
  "Species"="Species",
  "NumberOfYearsSeen"="Number Of Years Seen")
ft
```

### Rare Birds
These rare bird species were seen on only one field trip.
Again, only birds seen by both scouts and the leaders are considered.
```{r rare_birds, echo=FALSE}
rare_birds <- sort(year_count[year_count < 2])
indx_rare <- match(names(rare_birds), birds_by_year[["Species"]])
year_seen <- birds_by_year[indx_rare, "Year"]
tdf <- data.frame(Species=names(rare_birds), YearSeen=as.character(year_seen))
ft <- flextable(tdf, cwidth=c(3.0, 2.0))
ft <- set_header_labels(ft,
  "Species"="Species",
  "YearSeen"="Year Seen")
ft <- align(ft, j=2, align = "right", part = "all")
ft
```

### All Birds
All bird species seen by year on all the field trips.
```{r birds_vs_years, fig.width=8, fig.height=9, echo=FALSE}
birds_by_year |>
  mutate(fYear = as.factor(Year), fSpecies = factor(Species, levels=rev(ordered_species))) |>
  ggplot(aes(fYear, fSpecies, color=SeenBy)) +
    geom_point() +
    xlab("Year") +
    ylab("Species") +
    theme_light() +
    theme(legend.position = "top")

```
